---
title: "Transport flows in Bristol"
author: "Mark Padgham"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: true
        toc_float: true
        number_sections: false
        theme: flatly
---

# Using dodgr to generate flows throughout the Bristol street network

Note at the outset that these routines likely require the latest, development
version of `dodgr`, installed with
```{r, eval = FALSE}
devtools::install_github ("ATFutures/dodgr")
```
and not the somewhat older CRAN version installed with
```{r, eval = FALSE}
install.packages ("dodgr")
```

## 1. Extract street network

Street networks and associated boundaries can be conveniently extracted from
Open Street Map with the [`osmdata`
package](https://github.com/ropensci/osmdata).  We first need to get a polygonal
boundary for Bristol
```{r osmdata, message = FALSE}
library (osmdata)
bp <- getbb (place_name = "Bristol UK", format_out = "polygon")
class (bp)
length (bp)
```
The returned object is a list with two polygons. The easiest way to work out
which one we need is to look at them:
```{r bristol-poly}
par (mfrow = c (1, 2))
plot (bp [[1]], type = "l")
plot (bp [[2]], type = "l")
```

The first one is obviously what we want (the second extents to the two islands
of Flat Holm and Steep Holm within the River Severn, which are offically part of
Bristol). The street network can be downloaded with `osmdata` using the
following lines
```{r bristol-extract, eval = FALSE}
net <- osmdata::opq (bp [[1]]) %>%
    osmdata::add_osm_feature (key = "highway") %>%
    osmdata::osmdata_sf (quiet = FALSE) %>%
    osmdata::osm_poly2line () %>%
    osmdata::trim_osmdata (bp [[1]])
```
```{r bristol-save, echo = FALSE, eval = FALSE}
save (net, file = "bristol-net.rda")
format (file.size ("bristol-net.rda"), big.mark = ",") # 48.5MB
```
```{r bristol-load, echo = FALSE}
load ("bristol-net.rda")
```
Each of those lines does the following:
1. `opq` converts the bounding polygon to a query to be submitted to the Open
   Street Map [`overpass` server](https://overpass-turbo.eu);
2. `add_osm_feature` says we want to extract objects of class `"highway"`;
3. `osmdata_sf` says we want the data in Simple Features (`sf`) format;
4. `osm_poly2line` is necessary because street components that are polygon are
   stored in a separate item to non-polygonal lines; this command merges the
   two; and finally
5. `trim_osmdata` trims the data to within our specified bounding polygon

The result is an object of class `osmdata`:
```{r bristol-data-print}
net
```
The `osm_poly2line` function merges the `osm_polygons` with the `osm_lines`
objects, so that all street segments are then contained in the single
`osm_lines` object:
```{r bristol-lines}
net <- net$osm_lines
dim (net)
```
The city has just under 20,000 streets, and there are 254 columns in the `sf
data.frame`. These contain all kinds of information on each of the streets:
```{r}
names (net)
```

The [`dodgr` package](https://github.com/ATFutures/dodgr) (**D**istances on
**D**irected **Gr**aphs) contains a helper function for extracting street
networks:
```{r, eval = FALSE}
library (dodgr)
net2 <- dodgr_streetnet ("Bristol UK")
dim (net2)
```
```{r, echo = FALSE}
library (dodgr)
net2 <- array (NA, dim = c (31401, 257))
dim (net2)
```
This function yields more streets because it only extracts rectangular street
networks. Direct use of the `osmdata` package allows finer control. From here
on, we'll use the first, polygonal network.

## 2. dodgr distances

[`dodgr`](https://github.com/ATFutures/dodgr) can be used to calculate distances
between points. Let's work out how far it is to walk from the main train station
(Temple Meads) to the mouth of the Avon (where there is a railway station called
"Avonmouth")
```{r get-station-avonmouth}
station <- osmdata::opq (bp [[1]]) %>%
    osmdata::add_osm_feature (key = "building") %>%
    osmdata::add_osm_feature (key = "name", value = "Temple Meads",
                              value_exact = FALSE) %>%
    osmdata::osmdata_sf ()
avonmouth <- osmdata::opq (bp [[1]]) %>%
    osmdata::add_osm_feature (key = "railway") %>%
    osmdata::add_osm_feature (key = "name", value = "Avonmouth") %>%
    osmdata::osmdata_sf ()
```
Those are two `osmdata` objects with the points we want contained in the
`osm_polygons` item. Let's just get the distance between the first point of each
polygon
```{r station-avonmouth-to-points}
station <- station$osm_polygons %>%
    sf::st_coordinates ()
station <- as.numeric (station [1, 1:2])
avonmouth <- avonmouth$osm_polygons %>%
    sf::st_coordinates ()
avonmouth <- as.numeric (avonmouth [1, 1:2])
station; avonmouth
```
The final step before we can route through the street network is to convert the
`sf`-format network into a `dodgr` network. This is necessary in order to weight
each street segment according to a specified weighting profile. In this case, we
want to walk, so,
```{r, weight-streetnet}
net_walk <- weight_streetnet (net, wt_profile = "foot")
dim (net_walk)
```
The `weight_streetnet` function breaks the network down into individual street
segments, of which there are 176,000. Each of these is allocated a weighted
distance dependeing on the kind of way and the preferred mode of transport.
Calculating the distances is then as easy as,
```{r dodgr-dists}
dodgr_dists (net_walk, from = station, to = avonmouth)
```

### 2.1 Street networks and routing

To find out how far it is to cycle to Avonmouth, we need to generate a new
`dodgr` street network weighted for bicycle travel.
```{r weight-bike}
net_bike <- weight_streetnet (net, wt_profile = "bicycle")
dodgr_dists (net_bike, from = station, to = avonmouth)
```
It is about half a kilometre shorter to cycle than to walk to Avonmouth,
presumably because Bristol has put some excellent cyclepaths somewhere along the
way. Note that routing different kinds of transport **always** requires entirely
different street networks, weighted for each particular kind. 

## 3 Routes and dodgr

Where do our routes actually go? We can find out with `dodgr_paths`:
```{r dodgr_paths}
p_foot <- dodgr_paths (net_walk, from = station, to = avonmouth)
class (p_foot)
length (p_foot)
head (p_foot [[1]] [[1]])
```
`dodgr_paths` returns lists because it's designed for many-to-many routing tasks
(see below), where the first list item is the `[[from]]` entry, and the second
is the `[[to]]` entry. Each of these then contains a sequence of Open Street Map
node IDs. How do we map these back on to the network to generate the
coordindates of this path?
```{r dodgr_verts}
verts <- dodgr_vertices (net_walk)
head (verts)
```
The `dodgr_vertices` function extracts all of the nodal vertices from the
network. Each of these has an `id` value, so we just need to match our path to
these:
```{r index-verts}
index <- match (p_foot [[1]] [[1]], verts$id)
p_foot <- verts [index, ]
```
We can view this via `mapview` (or any other suitable way) by converting this to
an `sf` object
```{r path-to-sf}
p_foot <- p_foot [, c ("x", "y")] %>%
    as.matrix () %>%
    sf::st_linestring () %>%
    sf::st_sfc ()
sf::st_crs (p_foot) <- 4326 # OSM CRS
```
Then we're good to go. The following code will open up an interactive map of our
path.
```{r, eval = FALSE}
library (mapview)
mapview (p_foot)
```
We can compare the two of them by adding the `bike` path in a different colour:
```{r bike-path, eval = FALSE}
p_bike <- dodgr_paths (net_bike, from = station, to = avonmouth)
verts <- dodgr_vertices (net_bike)
index <- match (p_bike [[1]] [[1]], verts$id)
p_bike <- verts [index, ]
p_bike <- p_bike [, c ("x", "y")] %>%
    as.matrix () %>%
    sf::st_linestring () %>%
    sf::st_sfc ()
sf::st_crs (p_bike) <- 4326 # OSM CRS
mapview (p_foot) %>%
    addFeatures (p_bike, color = "red")
```
Running this code will reveal why and precisely where the walking path is
slightly longer than the cycling path.

### 3.1 What is the advantage of dodgr?

Thus far, [`dodgr`](https://github.com/ATFutures/dodgr) does nothing more than
what [`stplanr`](https://github.com/ropensci/stplanr) can do, yet takes
considerably more code to get there. The real advantage of `dodgr`, and the
purpose of its development, is in **massive** routing tasks. `dodgr` can
calculate distances between loads of points really quickly. Let's take a sample
of the vertices listed previously (and it suffices to just pass the OSM `"id"`
values to these `dodgr` routines):
```{r}
vert_sample <- verts$id [sample (nrow (verts), size = 100)]
system.time (
dists <- dodgr_dists (net_bike, from = vert_sample, to = vert_sample)
)
```
That call just calculated $100\times 100=10,000$ distances in around one second.
Such calculations can technically be done by calling routing APIs, but even
where that is possible, the big constraint is the number of calls. google, for
example, only offers 2,500 free requests per day, so we're already over the
limit there. They then charge US\$0.50 per additional 1,000 requests, so that
would have cost us $7.5\times 0.5 = \$US3.25$. `dodgr` can scale to far larger
calls. Try scaling the previous call to 1,000 vertices instead of 100. That
generally takes around one minute, and performs one million calculations. Using
the google API, that would translate to $1,000\times 0.5 = \$US500$.

#### Advantage #1: dodgr is free!

Not only is `dodgr` free, it is also very highly optimized code. Let's compare
`dodgr` with the "routing-industry-standard" [`igraph`](http://igraph.org).
We'll do the comparison on a smaller sub-sample of the graph, extracted using
the `dodgr_sample` function. We also need to do some fiddling around to convert
our network to `igraph` format.
```{r igraph-benchmark}
nets <- dodgr_sample (net_bike, nverts = 1000)
netc <- dodgr_contract_graph (nets)$graph # removes all points except junctions
from_id <- unique (netc$from_id)
to_id <- unique (netc$to_id)

# set up igraph:
edges <- cbind (netc$from_id, netc$to_id)
edges <- as.vector (t (edges))
igr <- igraph::make_directed_graph (edges)
igraph::E (igr)$weight <- netc$d_weighted

rbenchmark::benchmark (
                       d <- dodgr_dists (netc, from = from_id, to = to_id),
                       d <- igraph::distances (igr, v = from_id, to = to_id,
                                               mode = "out"),
                       replications = 10, order = "relative")
```

#### Advantage #2: dodgr is fast!

`dodgr` is a few **times** faster than the faster alternative. In computing
times, speed increases of a few percent are often big news. A few **times**
really is an awful lot faster. (Note also that `igraph` does some clever caching
behind the scenes, so re-running the above benchmark code will greatly increase
the relative speed of `igraph`; fair comparisons require re-running in a fresh
**R** session.)

## 4 dodgr and aggregated flows 

Beyond the usual distances and routes, the real workhorse of
[`dodgr`](https://github.com/ATFutures/dodgr) is the ability to aggregate flows
between multiple points. This can readily be demonstrated using our Bristol
network, calculating flows between a random sample of 10 points. Flow
aggregation requires specifying a `"flow"` matrix, equivalent to an OD matrix.

```{r flows}
n <- 10
pts <- sample (verts$id, size = n)
flows <- array (runif (n ^ 2), dim = rep (n, 2)) # random OD values
f <- dodgr_flows_aggregate (net_bike, from = pts, to = pts, flows = flows)
head (f)
```
The `dodgr_flows_aggregate()` function returns the input data (`net_bike`)
modified through the additional of an extra `flow` column. Plotting these flows
is best done through first filtering out all `flow == 0` values:
```{r}
f <- f [which (f$flow > 0), ]
```
This `dodgr data.frame` can then be converted to an `sf` object for plotting
with
```{r, eval = FALSE}
fsf <- dodgr_to_sf (f)
```
This is not quite a standard `sf` object, as the two components of data and
geometry are split into separate list items. The following lines can be used to
plot the results, scaling both line thicknesses and colours according to the
flow values.
```{r, eval = FALSE}
cols <- topo.colors (30) [ceiling (30 * fsf$dat$flow / max (fsf$dat$flow))]
mapview (fsf$geoms, lwd = fsf$dat$flow, color = cols)
```
That result is only from a made-up sample of 10 points. It is obviously
straightforward both to extend to many more points, and also to insert real OD
data. As a final note, `mapview` - and any other `leaflet`-based map generators
- struggles with large data sets (say, beyond 1,000 or so `sf` objects). For
such large data sets, it is often advantageous to filter out flows below some
finite minimal value, through replacing the above line with
```{r, eval = FALSE}
flim <- 0.1 # or whatever; examine histogram of flows?
f <- f [which (f$flow > flim), ]
```

### 4.1 Dispersed flows

Finally, [`dodgr`](https://github.com/ATFutures/dodgr) includes an additional
function, `dodgr_flows_disperse`. This calculates flows describing movement that
simply disperses away from a given set of points according to a simple
exponential dispersal model. This model corresponds to a typical spatial
interaction model, and this function can be applied in places lacking OD data
but for which population densities are available.
```{r, eval = FALSE}
f <- dodgr_flows_disperse (net_bike, from = pts, dens = runif (length (pts)),
                           k = 0.5) # k is exponential decay parameter in km
```
This can also be plotted with the same code as above, although (i) it may not
respond as many more line segments will be generated, and (ii) the line
thicknesses may need to be adjusted with
```{r, eval = FALSE}
line_scale <- 2 # or whatever
mapview (fsf$geoms, lwd = line_scale * fsf$dat$flow, color = cols)
```
